---
title: "HW 3"
author: "Carolyn Martinez"
date: "2024-10-10"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r loading libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## Problem 1: NY NOAA Data 

```{r loading and claening data}

library(p8105.datasets)
data("ny_noaa")
```

The data has observations on `r nrow(ny_noaa)` individual GHCND-Daily weather stations. The data set has `r ncol(ny_noaa)` variables some of which include the station ID, the date of the observation, the precipitation in tenths of a mm, the snowfall in mm, the snow depth in mm, and the max and temperature in degrees Celsius. Missing data is a big issue considering there are `r sum(is.na(ny_noaa))` missing data in the set. 

```{r this is separating date & changing units}

ny_noaa |>
   na.omit()|>
separate(date, into = c("year", "month", "day"), sep="-", extra = "merge", fill = "right")|>
   mutate(
      prcp = round(as.numeric(prcp, digits =0)),
      tmin = as.numeric(tmin)/10,
      tmax = as.numeric(tmax)/10,
      snow = as.numeric(snow)
      )|>
   tabulate(unique(snow))|>
   which.max()
   
```






